Multitask matrix completion for learning protein interactions across diseases

### 2016-08-21 paper.tex
first commit

\documentclass[runningheads,a4paper]{llncs}
%\usepackage[top=1.3in, bottom=1.3in, left=1.3in, right=1.3in]{geometry}
\usepackage{floatrow}
\usepackage{graphicx} % more modern
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}
\usepackage{subfigure} 
%\usepackage{natbib}
\usepackage[footnotesize]{caption}
\usepackage{multirow}
\usepackage{makeidx}

@@

\newcommand{\specialcell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\usepackage{url}
\urldef{\mailsa}\path|mkshirs@us.ibm.com|
\urldef{\mailsb}\path|{jgc,judithks,kmuruges}@cs.cmu.edu|
%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%\setcounter{secnumdepth}{0}

\begin{document}
\pagestyle{headings}

@@

\author{Meghana Kshirsagar$^{1}$%
\and Jaime G. Carbonell$^{2}$
\and Judith Klein-Seetharaman$^{3}$
\and Keerthiram Murugesan$^{2}$}

@@

\institute{$^{1}$IBM T. J. Watson Research, Yorktown Heights, NY 10598, USA \\
$^{2}$Language Technologies Institute, Carnegie Mellon University, 5000 Forbes Ave., Pittsburgh PA 15213, USA\\
$^{3}$Metabolic \& Vascular Health, Warwick Medical School, University of Warwick, Coventry, UK\\
\mailsa \quad \mailsb
}

@@

Disease causing pathogens such as viruses, introduce their proteins into the host cells 
where they interact with the host's proteins enabling the virus to replicate inside the host.
These interactions between pathogen and host proteins are key to understanding infectious diseases.
Often multiple diseases involve phylogenetically related or biologically similar pathogens. Here we present a multitask learning method to jointly model interactions between human proteins and three different, but related viruses: \textit{Hepatitis C}, \textit{Ebola} virus and \textit{Influenza A}. 
Our multitask matrix completion based model uses a shared low-rank structure in addition to a task-specific sparse structure to incorporate the various interactions.
We obtain upto 30 \% improvement in predictive performance over prior state-of-the-art models. 
We show how our model's parameters can be interpreted to reveal both general and specific
interaction-relevant characteristics of the viruses. Our code and data is available at: \url{http://www.cs.cmu.edu/~mkshirsa/bsl_mtl.tgz}
\keywords{host-pathogen protein interactions, multitask learning, matrix completion}
\end{abstract}

@@

An elegant way to formulate the interaction prediction problem is via a graph completion based framework, where we have several bipartite 
graphs over multiple hosts and pathogens.% as illustrated in Figure \ref{fig:hpppigraph}.
Nodes in the graphs represent host and pathogen proteins, with edges between them representing interactions 
(host protein $\frac{interacts}{}$ pathogen protein).
Given some observed edges (interactions obtained from laboratory based experiments), we wish to predict the other edges in the graphs.
Such bipartite graphs arise in a plethora of problems including: recommendation systems (user $\frac{prefers}{}$ movie), 
citation networks (author $\frac{cites}{}$ paper), disease-gene networks (gene $\frac{influences}{}$ disease) etc. 
In our problem, each bipartite graph $\mathcal{G}$ can be represented using a matrix $M$, where the rows correspond to pathogen proteins 
and columns correspond to host proteins.
The matrix entry $M_{ij}$ encodes the edge between pathogen protein $i$ and host protein $j$ from the graph, with $M_{ij}=1$ for the observed interactions. 
Thus, the graph completion problem can be mathematically modeled as a matrix completion problem \cite{candes08}. 
%Traditional approaches towards matrix completion rely on the assumption that the underlying function that generated the matrix can be decomposed into a small number of `latent' factors. The solution based on this assumption involves finding a low-rank matrix factorization for $M$, mathematically expressed as: finding $U$ and $V$ such that $M \approx UV^T$. Here the parameters $U$ and $V$ are called the factor matrices and represent the latent properties of the host and pathogen proteins respectively. In the recommendation systems example, the `low rank' structure suggests that movies can be grouped into a small number of latent `genres'. In the case of proteins these latent properties could correspond to various \textit{biological functions} of proteins or encode \textit{interaction propensities}.

Most of the prior work on host-pathogen PPI prediction has modeled each bipartite graph separately, and hence cannot exploit the similarities 
in the edges across the various graphs. Here we present a \textit{multitask} matrix completion method that \textit{jointly models} 
several bipartite graphs by sharing information across them. 
From the multitask perspective, a \textit{task} is the graph between one host and one pathogen (can also be seen as interactions relevant to one disease). 
We focus on the setting where we have a single host species (human) and several related viruses, where we hope to gain from the 
fact that similar viruses will have similar strategies to infect and hijack biological processes in the human body.
%Such opportunities for sharing arise in other applications as well: for instance, predicting user preferences in movies may inform preferences in selection of books, or vice-versa, as movies and books are semantically related. Multitask learning based models that incorporate and exploit these correlations should benefit from the additional information.
% and result in a better prediction performance. 
Our model is motivated by the following biological intuition governing protein interactions across diseases.
\begin{enumerate}
\item An interaction depends on the structural properties of the proteins, which are conserved across similar viruses as they have evolved from common ancestors. We use a component to capture these latent similarities, that is \textit{shared} across tasks.
%This latent structural similarity can be expressed via latent factors that project the proteins into a shared lower dimensional subspace that is common to all tasks.
\item In addition to the shared properties discussed above, each pathogen has also evolved specialized mechanisms to target host proteins. These are unique to the pathogen and can be expressed using a \textit{task-specific} parameter.
\end{enumerate}

@@

\begin{table}
\caption{Tasks and their sizes. Each column corresponds to one bipartite graph between human proteins and the pathogen indicated in the column header. All pathogens are single stranded RNA viruses. The interactions and the protein count both includes data across various strains of each pathogen}
% (see Sec \ref{sec:datasets} for details). The last row shows that each of our graphs is extremely sparse.} 
\label{datasets}
\begin{footnotesize}
\begin{tabular}{crrr} \toprule
Task $\rightarrow$ & \textbf{\textit{Influenza A}} & \textbf{\textit{Hepatitis C}} & \textbf{\textit{Ebola}}\\ \midrule 
number of HP PPIs (positives) & 848& 981 & 90 \\\hline
\# of distinct virus proteins in PPIs & 54 & 151 & 2 \\\hline
{\# of distinct human proteins in PPIs} & 362 & 385 & 88 \\\hline
{total \# of virus proteins across strains} & 542 & 163 & 150 \\\hline
{number of negatives} & 84800 & 98100 & 9000 \\\hline
{density of observed graph$^\ddagger$ (as \%)} & .15 & .60 & .06 \\\hline
\multicolumn{4}{l}{\scriptsize{HP PPI: host-pathogen protein protein interactions}}\\
\multicolumn{4}{l}{\scriptsize{$\ddagger$: considering all proteins from the two tasks involved. }}\\
\multicolumn{4}{l}{\scriptsize{Note: considering the total number of human proteins to be $\approx$ 100,000.}}\\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table}

To incorporate the above ideas, we assume that the interactions matrix $M$ is generated from two components. 
The first component has low-rank latent factors over the human and virus proteins,
with these latent factors jointly learned over all tasks. The second component involves a task specific parameter, on which we
additionally impose a sparsity constraint as we do not want this parameter to overfit the data. Section \ref{ourmodel} discusses
our model in detail. We trade-off the relative importance of the two components using task-specific hyperparameters.
We can thus learn what is conserved and what is different across pathogens, rather than having to specify it manually.

%One of the key challenges in inducing such a model is: in addition to the graph structure itself, we want to incorporate information about the nodes available in the form of features. 
The applications that we consider involve extremely sparse graphs with a large number of nodes and very few observed edges. There will be nodes i.e proteins that are not involved in any known interactions -- the model should be able to predict links between such prior `unseen' node pairs (called the \textit{cold start problem} in the recommendation systems community). 
For instance, the host-pathogen PPI network of human-Ebola virus (column-3, Table \ref{datasets}) has $\approx$ 90 observed edges (equivalent to 0.06\% of the possible edges) which involve only 2 distinct virus proteins. Any biologist studying virus-human interactions will be more interested in the virus proteins which have yet unknown interactions.
The main contributions of this work are:\\
1. We extend a prior matrix completion model \cite{abernethy} to the multitask setting.\\
2. We leverage node features which allows us to predict on unseen nodes.\\
3. We apply the model to an important problem -- prediction of interactions in disease-relevant host-pathogen protein networks, for multiple related diseases and demonstrate improved performance over prior state-of-the-art multitask models.\\
4. We use unlabeled data to initialize the parameters of our model, which gives us a modest boost in prediction performance.

\subsection{Background: Host-pathogen protein interactions}
%An important part of our sustenance depends on how we counter infectious diseases -- those that are caused by external agents like pathogens. To understand how these microbial organisms reside and disrupt the host processes we need to comprehend the cross-organism protein interactions between the host species such as mammals, and the causative pathogen species such as viruses and bacteria. %These interactions  pathogen proteins physically bind with human proteins to manipulate important biological processes in the host cell, evade the host immune response and multiply within the host. 
The experimental discovery of host-pathogen protein interactions involves biochemical and biophysical methods such as co-immunoprecipitation (co-IP), yeast two-hybrid (Y2H) assays, co-crystalization. % These databases are valuable sources of information to bioinformaticians for developing models. \\
The most reliable experimental methods are often very time-consuming and expensive, making it hard to investigate the prohibitively large set of possible host-pathogen interactions -- e.g. the bacterium \textit{Bacillus anthracis} with about 2321 proteins when coupled with the 100,000 human proteins gives $\approx$232 million protein pairs to validate. Computational techniques complement laboratory-based methods by predicting highly probable PPIs. Supervised machine learning based methods use the few known interactions as training data and formulate the interaction prediction problem in a classification setting%, with target classes: ``interacting'' or ``non-interacting''. Features are derived using various attributes of the two proteins. %such as: protein sequences from Uniprot \cite{uniprot}, protein structure from PDB, gene ontology from GO database \cite{GO}. 
%In our work we use protein-sequence based features.

\subsection{Prior work}
%\label{prior_work}
Most of the prior work in PPI prediction has focussed on building models separately for individual 
organisms \cite{struct2net,qi06} or on building a model specific to a disease in the case of host-pathogen PPI 
prediction \cite{tastan09,dyer07,me12}. There has been limited work on combining PPI datasets to learn joint models. \cite{qi2010} proposed a semi-supervised multi-task framework to predict PPIs from partially labeled reference sets. \cite{me_ismb_2013} develop a task regularization based
framework that incorporates the similarity in biological pathways targeted by various diseases.
%  to couple multiple tasks together. %Matrix factorization based protein-protein interaction (PPI) prediction has seen very little work, mainly due to the extremely sparse nature of these datasets which makes it very difficult to get reliable predictors. 
\cite{ppi_cmf} uses a collective matrix factorization based approach in a multi-task learning setting for within species PPI prediction. The methods used 
in all prior work on PPI prediction do not explicitly model the features of the proteins and cannot be applied on proteins which have no known interactions available. Our work addresses both these issues.
%in using a formulation of the matrix completion problem originally proposed in the work by \cite{abernethy}.
%A majority of the prior work in the relevant areas of collaborative filtering and link prediction includes single relation models that use neighbourhood based prediction \cite{sarwar01}, matrix factorization based approaches \cite{koren09,menon_ecml11} and bayesian approaches using graphical models \cite{jin02,phung09}. 
%The matrix factorization based methods have been more popular than the other methods. 
%There have also been multitask approaches on link prediction \cite{zhang12,cao_icml2010,li09,cmf}. 
%\cite{li11} presents a survey on multitask/transfer methods in collaborative filtering. 
%The multi-relational learning literature \cite{xu09} and the work on link prediction in heterogenous networks is not relevant as their setting involves \textit{different types of relationships} between the \textit{same} set of nodes.
%The prior work on incorporating side information in the form of features of the nodes (users/movies) and edges
%has been proposed for single-graph link prediction problems. 
%\cite{menon_ecml11} propose a single-graph model that combines linear and bilinear features, latent parameters on the nodes and several other parameters into a function that minimizes a ranking loss. 
%Probabilistic models to combine network structure with node features have also been proposed. 
We extend the framework from \cite{abernethy} to the multitask setting. 
%There has been a lot of work on other low-rank models for multitask learning \cite{ando2005,ji2009,chen2012,chen2013}. 
% that try to capture the task relationship via a shared low-rank structure over the model parameters. Chen et al \cite{chen2012} have developed efficient algorithms for sparse and low-rank multitask learning models for classification.

@@

\subsection{Handling the `curse of missing negatives'}
\label{negs}
For the MC algorithm to work in practice the matrix entries $M_{ij}$ should represent interaction scores (range [0 1]) 
or take binary values (1s for positives and 0s for negatives). Our experiments with PPI probabilities (obtained using the MINT-scoring algorithm)
gave bad models. The binary matrix setting requires some observed 0s. However non-interactions are not available as they cannot be verified 
experimentally for various reasons. Hence we derived a set of `probable negatives' using a heuristic often used in PPI prediction 
work \cite{qi06,qi09,dyer11,me_ismb_2013}. We pair up all virus proteins with all human proteins and sample a random set to be negatives. 
This heuristic works in practice as the interaction ratio (i.e number of positives in a large random set of protein pairs) is expected to be very low:
$\approx$ 1/100 to 1/500. That is, the probability that our negatives contain true positives is negligible. \\
\textbf{High class imbalance}: We incorporate the prior on the interaction ratio by setting the size of our randomly sampled negatives set equal to
100 times the number of gold standard positives.

\section{Experimental setup}
\subsection{Datasets and features}
\label{sec:datasets}
We use three human-virus PPI datasets from the PHISTO \cite{phisto} database (version from 2014), the characteristics of which are summarized in Table \ref{datasets}. 
The \textit{Influenza A} task includes various strains of flu: H1N1, H3N2, H5N1, H7N3. 
Similarly, the \textit{Hepatitis} task includes various strains of the virus. \footnote{Since we use data from several strains for each task, the PPI data contains some interactions that are interologs. Please see the supplementary for details} 
All three are single-strand RNA viruses, with \textit{Hepatitis} being a positive-strand ssRNA whereas \textit{Influenza} and \textit{Ebola} are negative-strand viruses. %Note that we use UniprotKB identifiers to refer to the proteins; hence if an identical protein appears in two strains it will be treated differently as it has a different UniprotKB id. 
The density of the known interactions is quite small when considering the entire proteome (i.e all known proteins) of the host and pathogen species (last row in Table \ref{datasets}).\\
\noindent\textbf{Features}: 
Since the sequence of a protein determines its structure and consequently its function, it may be possible to predict PPIs using the amino acid sequence of a protein pair. \cite{shen:2007} introduced the ``conjoint triad model'' for predicting PPIs using only amino acid sequences. They partitioned the twenty amino acids into seven classes based on their electrostatic and water affinities.\footnote{For details of these classes, please refer to the supplementary or the original paper}
A protein's amino acid sequence is first transformed to a class-sequence (by replacing each amino acid by its class).
For $k$=3, they count the number of times each distinct tri-mer (set of three consecutive amino acids) occurred in the sequence.
Since there are 343 ($7^3$) possible tri-mers (with an alphabet of size 7), the feature vector containing the tri-mer frequency counts will have 343 elements.
To account for protein size, they normalized the counts by linearly transforming them to lie between 0 and 1.
Thus the value of each feature in the feature vector is the normalized count for each of the possible amino acid three-mers. We use di-, tri- and four-mers thus leading to a total of 2793 features ($7^2 + 7^3 + 7^4$).
Such features have been successfully applied in prior work \cite{dyer07,me_ismb_2013}.

\subsection{Competing methods}
The results upon comparing our model to various single and multitask methods are shown in Table \ref{resultsTable}: including other recent low-rank and sparse models, conventional multitask methods and prior work on HP PPI prediction. Wherever appropriate, we concatenated the features of the two node types into a single feature vector. Let $W \in \mathbb{R}^{T \times d_t}$ be the matrix with the task-specific weight vectors $\mathbf{w}_t$. For a uniform comparison we used least squared loss in all the methods. The MALSAR package was used for some of the baselines.

>>>>

\documentclass[runningheads,a4paper]{llncs}
\usepackage[top=1.3in, bottom=1.3in, left=1in, right=1in]{geometry}
\usepackage{floatrow}
\usepackage{graphicx} % more modern
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}
\usepackage{subfigure} 
\usepackage{natbib}
\usepackage[footnotesize]{caption}
\usepackage{multirow}

@@

\usepackage{url}
\urldef{\mailsa}\path|{mkshirsa,jgc,judithks,kmuruges}@cs.cmu.edu|
%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\setcounter{secnumdepth}{0}

\begin{document}

@@

\author{Meghana Kshirsagar$^{1}$%
\and Jaime G. Carbonell$^{1}$
\and Judith Klein-Seetharaman$^{2}$
\and Keerthiram Murugesan$^{1}$}

@@

\institute{$^{1}$Language Technologies Institute, Carnegie Mellon Univ., 5000 Forbes Ave., Pittsburgh PA 15213, USA\\
$^{2}$Metabolic \& Vascular Health, Warwick Medical School, Univ. of Warwick, Coventry, UK\\
\mailsa
}

@@

Disease causing pathogens such as viruses, introduce their proteins into the host cells 
where they interact with the host's proteins enabling the virus to replicate inside the host.
These interactions between pathogen and host proteins are key to understanding infectious diseases.
Often multiple diseases involve phylogenetically related or biologically similar pathogens. Here we present a multitask learning method to jointly model interactions between human proteins and three different, but related viruses: \textit{Hepatitis C}, \textit{Ebola} virus and \textit{Influenza A}. 
Our multitask matrix completion based model uses a shared low-rank structure in addition to a task-specific sparse structure to incorporate the various interactions.
We obtain between 7 to 39 \% points improvement in predictive performance over prior state-of-the-art models. 
We show how our model's parameters can be interpreted to reveal both general and specific
interaction-relevant characteristics of the viruses. Our code and data is available at: \url{http://www.cs.cmu.edu/~mkshirsa/bsl_mtl.tgz}
\keywords{host-pathogen protein interactions, multitask learning, matrix completion}
\end{abstract}

@@

An elegant way to formulate the interaction prediction problem is via a graph completion based framework, where we have several bipartite 
graphs over multiple hosts and pathogens as illustrated in Figure \ref{fig:hpppigraph}.
Nodes in the graphs represent host proteins (circles) and pathogen proteins (triangles), with edges between them representing interactions 
(host protein $\frac{interacts}{}$ pathogen protein).
Given some observed edges (interactions obtained from laboratory based experiments), we wish to predict the other edges in the graphs.
Such bipartite graphs arise in a plethora of problems including: recommendation systems (user $\frac{prefers}{}$ movie), 
citation networks (author $\frac{cites}{}$ paper), disease-gene networks (gene $\frac{influences}{}$ disease) etc. 
In our problem, each bipartite graph $\mathcal{G}$ can be represented using a matrix $M$, where the rows correspond to pathogen proteins 
and columns correspond to host proteins.
The matrix entry $M_{ij}$ encodes the edge between pathogen protein $i$ and host protein $j$ from the graph, with $M_{ij}=1$ for the observed interactions. 
Thus, the graph completion problem can be mathematically modeled as a matrix completion problem \citep{candes08}. 
%Traditional approaches towards matrix completion rely on the assumption that the underlying function that generated the matrix can be decomposed into a small number of `latent' factors. The solution based on this assumption involves finding a low-rank matrix factorization for $M$, mathematically expressed as: finding $U$ and $V$ such that $M \approx UV^T$. Here the parameters $U$ and $V$ are called the factor matrices and represent the latent properties of the host and pathogen proteins respectively. In the recommendation systems example, the `low rank' structure suggests that movies can be grouped into a small number of latent `genres'. In the case of proteins these latent properties could correspond to various \textit{biological functions} of proteins or encode \textit{interaction propensities}.

Most of the prior work on host-pathogen PPI prediction has modeled each bipartite graph separately, and hence cannot exploit the similarities 
in the edges across the various graphs. Here we present a \textit{multitask} matrix completion method that \textit{jointly models} 
several bipartite graphs by sharing information across them. 
From the multitask perspective, a \textit{task} is the graph between one host and one pathogen (can also be seen as interactions relevant to one disease). 
We focus on the setting where we have a single host species (human) and several related viruses, where we hope to gain from the 
fact that similar viruses will have similar strategies to infect and hijack biological processes in the human body.
Such opportunities for sharing arise in other applications as well: for instance, predicting user preferences in movies may inform preferences in selection of books, 
or vice-versa, as movies and books are semantically related. Multitask learning based models that incorporate and exploit these correlations should benefit from 
the additional information.% and result in a better prediction performance.

Our multitask matrix completion based model is motivated by the following biological intuition governing protein interactions across diseases.
\begin{enumerate}
\item An interaction depends on the structural properties of the proteins, which are conserved across similar viruses as they have evolved from common ancestors. Our model thus needs a component to capture these latent similarities, which is \textit{shared} across tasks.
%This latent structural similarity can be expressed via latent factors that project the proteins into a shared lower dimensional subspace that is common to all tasks.
\item In addition to the shared properties discussed above, each pathogen has also evolved specialized mechanisms to target host proteins. These are unique to the pathogen and can be expressed using a \textit{task-specific} parameter in the model.
\end{enumerate}

@@

\begin{figure*}
\begin{floatrow}
\ffigbox[0.29\textwidth]{%
%trim option's parameter order: left bottom right top
\includegraphics[scale=0.33, trim = 0.2 9.2cm 4cm 0]{fig1.eps}
}
{
\caption{Multiple bipartite graphs: on the left are host proteins and on the right are proteins from different virus species. Edges represent protein interactions.}
\label{fig:hpppigraph}
}
\capbtabbox[0.69\textwidth]{%
\begin{footnotesize}
\begin{tabular}{crrr} \toprule
Task $\rightarrow$ & \textbf{\textit{Influenza A}} & \textbf{\textit{Hepatitis C}} & \textbf{\textit{Ebola}}\\ \midrule 
number of HP PPIs (positives) & 848& 981 & 90 \\\hline
\# of distinct virus proteins in PPIs & 54 & 151 & 2 \\\hline
{\# of distinct human proteins in PPIs} & 362 & 385 & 88 \\\hline
{total \# of virus proteins across strains} & 542 & 163 & 150 \\\hline
{number of negatives} & 84800 & 98100 & 9000 \\\hline
{density of observed graph$^\ddagger$ (as \%)} & .15 & .60 & .06 \\\hline
\multicolumn{4}{l}{\scriptsize{HP PPI: host-pathogen protein protein interactions}}\\
\multicolumn{4}{l}{\scriptsize{$\ddagger$: considering all proteins in the two organisms involved. }}\\
\multicolumn{4}{l}{\scriptsize{Note: considering the total number of human proteins to be $\approx$ 100,000.}}\\
\bottomrule
\end{tabular}
\end{footnotesize}
}
{
\caption{Tasks and their sizes. Each column corresponds to one bipartite graph between human proteins and the pathogen indicated in the column header. All pathogens are single stranded RNA viruses. Row 4 shows that each of our graphs is extremely sparse.} 
\label{datasets}
}
\end{floatrow}
\end{figure*}

To incorporate the above ideas, we assume that the interactions matrix $M$ is generated from two components. 
The first component has low-rank latent factors over the human and virus proteins,
with these latent factors jointly learned over all tasks. The second component involves a task specific parameter, on which we
additionally impose a sparsity constraint as we do not want this parameter to overfit the data. Section \ref{ourmodel} discusses
our model in detail. We trade-off the relative importance of the two components using task-specific hyperparameters.
Our model can thus learn what is conserved and what is different across pathogens, rather than having to specify it manually.

The key challenges in inducing such a model are: (1) In addition to the interactions from each graph, it should exploit information available in the form of features. 
(2) Exploiting features is particularly crucial since the graph $\mathcal{G}$ is often extremely sparse, i.e there are a large number of nodes and very few edges are observed. 
There will be proteins (i.e nodes) that are not involved in any known interactions -- called the \textit{cold start problem} in the recommendation systems community. 
The model should be able to predict the existence of links (or their absence) between such prior `unseen' node pairs. 
This is of particular significance in graphs that capture biological phenomena. For instance, the host-pathogen PPI network
of human-Ebola virus (column-3, Table \ref{datasets}) has $\approx$ 90 observed edges (equivalent to 0.06\% of the possible edges) which involve only 2 distinct virus proteins. Any biologist studying virus-human interactions will be interested in knowing more about the hundreds of other virus proteins (which have yet unknown interactions).
(3) A side-effect of having scarce data is the availability of a large number of unlabeled examples, 
i.e pairs of nodes with no edge between them. 
These unlabeled examples can contain information about the graph as a whole, and a good model should be able to use them.
%(4) ---- class skew ----

The main contributions of this work are:
\begin{itemize}
\item We extend a prior matrix completion model \citep{abernethy} to the multitask setting. This extension is new.
\item Unlike most prior approaches, our model exploits node-based features which allows us to deal with the `cold start' problem (generating predictions on unseen nodes).
\item We apply the model to an important, real-world problem -- prediction of interactions in disease-relevant host-pathogen protein networks, for multiple related diseases. We demonstrate the superior performance of our model over prior state-of-the-art multitask models.
\item We use unlabeled data to initialize the parameters of our model, which serves as a prior. This gives us a modest boost in prediction performance.
\end{itemize}

\subsection{Background: Host-pathogen protein interactions}
%An important part of our sustenance depends on how we counter infectious diseases -- those that are caused by external agents like pathogens. To understand how these microbial organisms reside and disrupt the host processes we need to comprehend the cross-organism protein interactions between the host species such as mammals, and the causative pathogen species such as viruses and bacteria. %These interactions  pathogen proteins physically bind with human proteins to manipulate important biological processes in the host cell, evade the host immune response and multiply within the host. 
The experimental discovery of host-pathogen protein interactions involves biochemical and biophysical methods such as co-immunoprecipitation (co-IP), yeast two-hybrid (Y2H) assays, co-crystalization. The host-pathogen protein interactions from several small-scale and high throughput experiments are aggregated by databases such as virus-MINT\cite{virusmint}, HPIDB \cite{hpidb}, PHISTO \cite{phisto} etc by literature curation. These databases are valuable sources of information to bioinformaticians for developing models. \\

\noindent\textbf{Prediction of host-pathogen PPIs}: The most reliable experimental methods for studying protein-protein interactions (PPI) are often very time-consuming and expensive, making it hard to investigate the prohibitively large set of possible host-pathogen interactions -- for example, the bacterium \textit{Bacillus anthracis} which causes anthrax has about 2321 proteins which when coupled with the 100,000 or so human proteins gives $\approx$232 million protein pairs to test, experimentally. Computational techniques complement laboratory-based methods by predicting highly probable PPIs. These techniques use the known interactions data from previous experiments and predict the most plausible new interactions. In particular, supervised machine learning based methods use the few known interactions as training data and formulate the interaction prediction problem in a classification setting, with target classes: ``interacting'' or ``non-interacting''. Features are derived using various attributes of the two proteins such as: protein sequences from Uniprot \cite{uniprot}, protein structure from PDB, gene ontology from GO database \cite{GO}.

\subsection{Prior work}
%\label{prior_work}
Most of the prior work in PPI prediction has focussed on building models separately for individual 
organisms \citep{chen,yeast,struct2net,qi06} or on building a model specific to a disease in the case of host-pathogen PPI 
prediction \citep{tastan09,qi09,dyer07,me12}. There has been little work on combining PPI datasets with the goal of improving 
prediction performance for multiple organisms. \cite{qi2010} proposed a semi-supervised multi-task framework to predict PPIs from partially labeled reference sets. \cite{me_ismb_2013} develop a task regularization based
framework called MTPL that incorporates the similarity in biological pathways targeted by various diseases to 
couple multiple tasks together. Matrix factorization based protein-protein interaction (PPI) prediction has seen very little work, mainly due to the extremely sparse nature of these datasets which makes it very difficult to get reliable predictors. \cite{ppi_cmf} use a CMF-based approach in a multi-task learning setting for within species PPI prediction. The methods used 
in all prior work on PPI prediction do not explicitly model the features of the proteins and cannot be applied on proteins which have no known interactions available. Our work addresses both these issues.
%in using a formulation of the matrix completion problem originally proposed in the work by \cite{abernethy}.

A majority of the prior work in the relevant areas of collaborative filtering and link prediction includes single relation models that use neighbourhood based prediction \citep{sarwar01}, matrix factorization based approaches \citep{koren09,menon_ecml11} and bayesian approaches using graphical models \citep{jin02,phung09}. 
%The matrix factorization based methods have been more popular than the other methods. 
There have also been multitask approaches on link prediction \citep{zhang12,cao_icml2010,li09,cmf}. 
%\cite{li11} presents a survey on multitask/transfer methods in collaborative filtering. 
%The multi-relational learning literature \cite{xu09} and the work on link prediction in heterogenous networks is not relevant as their setting involves \textit{different types of relationships} between the \textit{same} set of nodes.
%The prior work on incorporating side information in the form of features of the nodes (users/movies) and edges
%has been proposed for single-graph link prediction problems. 
\cite{menon_ecml11} propose a single-graph model that combines linear 
and bilinear features, latent parameters on the nodes and several other parameters into a function that minimizes a ranking loss. 
%Probabilistic models to combine network structure with node features have also been proposed. 
\cite{abernethy} cast the problem of matrix completion in terms of the abstract problem of learning linear operators. Their framework allows the incorporation of features and kernels. We extend their bilinear model for the multitask setting. There has been a lot of work on other low-rank models for multitask learning \citep{ando2005,ji2009,chen2012,chen2013}. % that try to capture the task relationship via a shared low-rank structure over the model parameters. Chen et al \cite{chen2012} have developed efficient algorithms for sparse and low-rank multitask learning models for classification.

@@

\subsection{Handling the `curse of missing negatives'}
\label{negs}
For the MC algorithm to work in practice the matrix entries $M_{ij}$ should represent interaction scores (range [0 1]) 
or take binary values (1s for positives and 0s for negatives). Our experiments with PPI probabilities (obtained using the MINT-scoring algorithm)
gave bad models. The binary matrix setting requires some observed 0s. However non-interactions are not available as they cannot be verified 
experimentally for various reasons. Hence we derived a set of `probable negatives' using a heuristic often used in PPI prediction 
work \citep{qi06,qi09,dyer11,me_ismb_2013}. We pair up all virus proteins with all human proteins and sample a random set to be negatives. 
This heuristic works in practice as the interaction ratio (i.e number of positives in a large random set of protein pairs) is expected to be very low:
$\approx$ 1/100 to 1/500. That is, the probability that our negatives contain true positives is negligible. \\
\textbf{High class imbalance}: We incorporate the prior on the interaction ratio by setting the size of our randomly sampled negatives set equal to
100 times the number of gold standard positives.

\section{Experimental setup}
\subsection{Datasets and features}
\label{sec:datasets}
We use three human-virus PPI datasets from the PHISTO \citep{phisto} database (version from 2014), the characteristics of which are summarized in Table \ref{datasets}. 
The \textit{Influenza A} task includes various strains of flu: H1N1, H3N2, H5N1, H7N3. 
Similarly, the \textit{Hepatitis} task includes various strains of the virus. 
All three are single-strand RNA viruses, with \textit{Hepatitis} being a positive-strand ssRNA whereas \textit{Influenza} and \textit{Ebola} are negative-strand viruses. Note that we use UniprotKB identifiers to refer to
the proteins; hence if an identical protein appears in two strains it will be treated differently as it has a different
UniprotKB id. The density of the known interactions is quite small when considering the entire proteome (i.e all known proteins) of the host and pathogen species (row-4 in Table \ref{datasets}). \\
\noindent\textbf{Features}: 
Since the sequence of a protein determines its structure and consequently its function, it may be possible to predict PPIs using the amino acid sequence of a protein pair. \cite{shen:2007} introduced the ``conjoint triad model'' for predicting PPIs using only amino acid sequences. They partitioned the twenty amino acids into seven classes based on their electrostatic and water affinities.\footnote{For details of these classes, please refer to the supplementary or the original paper}
A protein's amino acid sequence is first transformed to a class-sequence (by replacing each amino acid by its class).
For $k$=3, they count the number of times each distinct tri-mer (set of three consecutive amino acids) occurred in the sequence.
Since there are 343 ($7^3$) possible tri-mers (with an alphabet of size 7), the feature vector containing the tri-mer frequency counts will have 343 elements.
To account for protein size, they normalized the counts by linearly transforming them to lie between 0 and 1.
Thus the value of each feature in the feature vector is the normalized count for each of the possible amino acid three-mers. We use di-, tri- and four-mers thus leading to a total of 2793 features ($7^2 + 7^3 + 7^4$).
Such features have been successfully applied in prior work \citep{dyer07,me_ismb_2013}.

\subsection{Competing methods}
The results upon comparing our model to various single and multitask methods are shown in Table \ref{resultsTable}. Our baselines include recent low-rank and sparse models, conventional multitask methods and prior work on HP PPI prediction. For a uniform comparison we used least squared loss in all the methods. The MALSAR package was used to implement some of the models. For the baselines wherever appropriate, we concatenated the features of the two node types into a single feature vector. Let $W \in \mathbb{R}^{T \times d_t}$ be the matrix with the task-specific weight vectors $\mathbf{w}_t$.

\noindent\textbf{Single task (STL)}: We used ridge regression with $\ell_2$ regularization (which performed better than $\ell_1$). \\
\noindent\textbf{MMTL}: The mean regularized multitask learning model from \cite{pontil04}. \\
\noindent\textbf{Dirty model} \citep{jalali2010}: This model assumes that $W = P + Q$, where $P$ enforces group sparsity and $Q$ is to control for element-wise sparsity. It uses the regularizer $\rho_1 \|P\|_{1,\infty} + \rho_2 \| Q \|_1$. \\
\noindent\textbf{Low rank model}: A low-rank structure is enforced on $W$ by minimizing the nuclear norm $\|W\|_*$. \\
\noindent\textbf{Sparse + low-rank} \citep{chen2012}: $W$ is assumed to have the decomposition: $W = P + Q$, where $P$ is sparse and $Q$ has a low-rank structure. \\
\noindent\textbf{IMC} \citep{prateek,nagarajan}: This is the link-prediction model from Section \ref{MCintro}, where data from all tasks is combined without incorporating any task relationships. $U$ and $V$ are shared by all tasks. We use the same initialization for this method as we do for our model. A comparison to this model tells us how much we gain from the task-specific sparsity component $S_t$. \\
\noindent\textbf{MTPL} \citep{me_ismb_2013}: A biologically inspired regularizer is used to capture task similarity.\\ \noindent\textbf{BSL-MTL}: Bilinear sparse low-rank multitask learning, the method developed in this paper.
